A comparison of design-based and model-based approaches for spatial data.

### 2021-09-03 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
incorporated revisions from Lisa.

\begin{figure}
\includegraphics[width=1\linewidth]{SpatialDVM_Manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. In the top row, we have one fixed population, and two random samples. In the bottom row, we have two realizations of the same spatial process sampled at the same locations.}\label{fig:fig1}
\end{figure}

@@

The design-based approach can use spatial locations to obtain spatially
balanced samples. First we discuss spatial balance with respect to the
population (Stevens Jr and Olsen, 2004). A sample is spatially balanced
with respect to the population if the sampled population units are a
miniature of the population units. A sample is a miniature of the
population if the distribution of the sampled population units mirrors
the density of all population units. Spatial balance with respect to the
population is different than spatial balance with respect to geography.
A sample that is spatially balanced with respect to geography is spread
out in some type of equidistant manner over geographical space and is
not meant to be miniatures of the population. When we refer to spatial
balance henceforth, we mean spatial balance with respect to the
population.

Spatially balanced samples are useful because they tend to yield
estimates that have lower variance than estimates constructed from
sampling designs lacking spatial balance (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens Jr and Olsen, 2004; Wang et al., 2013). To quantify
spatial balance, Stevens Jr and Olsen (2004) proposed loss functions
based on Voroni polygons. The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (Stevens Jr and Olsen, 2004). Since GRTS was
developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning (Robertson et al., 2018). We focus on the
Generalized Random Tessellation Stratified (GRTS) algorithm to select
spatially balanced sampling because it has several attractive properties
detailed by Stevens Jr and Olsen (2004) and Dumelle et al. (2021).

The GRTS algorithm is used to sample from finite and infinite
populations and works by utilizing a mapping between two-dimensional and
one-dimensional space. The population units in two-dimensional space are
divided into cells using a hierarchical index. Population units are then
mapped to a one-dimensional line via the hierarchical indexing. The line
length of each population unit equals its inclusion probability. A
systematic sample is conducted on the line and these samples are linked
to a population unit in two-dimensional space, which results in the
desired sample. Stevens Jr and Olsen (2004) provide and Dumelle et al.
(2021) provide further details.

After collecting a sample using the GRTS algorithm, the data are used to
estimate population parameters. The Horvitz-Thompson estimator (Horvitz
and Thompson, 1952) yields unbiased estimates of population means and
totals. For example, if \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\) (denoted by \(\hat{\tau}_{ht}\)),
is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) and \(\pi_i\) are the observed value and
inclusion probability of the \(i\)th population unit selected in the
sample. A similar formula exists for estimating the mean, \(\mu\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but they have two drawbacks. First, they rely
on calculating \(\pi_{ij}\), the probability that population unit \(i\)
and population unit \(j\) are included in the sample, and this can be
very difficult to calculate. Second, they ignore the spatial locations
of the population units. To address these drawbacks, Stevens Jr and
Olsen (2003) proposed a local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\), and it
incorporates spatial locations by assigning higher weights to nearby
observations. Stevens Jr and Olsen (2003) show this variance estimators
tends to reduce the estimated standard error of \(\hat{\tau}\), yielding
narrower confidence confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of basing inference off of a specific
sampling design, we assume the data are generated by a spatial process.
Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. Let
\(\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}\)
be a response variable that can be measured at the \(N\) population
units and is represented as an \(N \times 1\) vector. Suppose we want to
predict some linear function of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where \(\mathbf{b}\)
is a \(1 \times N\) vector of weights. For example, if we want to
predict the population total across all population units, then we would
use a vector of 1's for the weights.

Typically, however, we only have a sample of the \(N\) population units.
Denoting quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with a subscript \emph{u},

@@

We also typically assume that there is spatial correlation in
\(\bm{\delta}\), which can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j^{th}\) entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align} where \(h_{i,j}\) is the distance between population units
\(i\) and \(j\), and \(\bm{\theta}\) is a vector of spatial covariance
parameters of the partial sill \(\theta_1\), the range \(\theta_2\), and
the nugget \(\theta_3\), and \(\mathbbm{1}\) is an indicator function.
However, any spatial covariance function could be used in the place of
the exponential, including functions that allow for non-stationarity or
anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

@@

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. However,
k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013),
random forest (Breiman, 2001), Bayesian models (Chan-Golston et al.,
2020), among others, can also be used to obtain predictions for a mean
or total from spatially correlated responses in a finite population
setting.

@@

  \begin{itemize}
  \tightlist
  \item
    normal: mean 0, variance 2
  \item
    lognormal: log scale mean 0, log scale variance 2
  \item
    lognormalbig: log scale mean 0, log scale variance 4
  \end{itemize}
\item
  sample sizes: \(n = 10, 50, 150; N = 900\)
\item
  layout: gridded vs random population locations
\end{itemize}

@@

In each simulation, a GRTS sample and an IRS sample were selected. Then
for the GRTS sample, the design-based approach using the local
neighborhood variance (Design GRTS) and a model-based approach were
applied (Model GRTS). Then for the IRS sample, the design-based approach
using the simple random sample variance (Design IRS) and a model-based
approach were applied (Model IRS).

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  We will now focus in a bit more on comparing Design-GRTS to
  Model-GRTS, the two best approaches for any reasonable sample size. In
  the independent error settings, the two approaches perform very
  similarly, so those results are omitted in the following graph. In the
  dependent error settings, using rmspe as the performance criterion,
  Model-GRTS outperforms Design-GRTS in 12 of the 18 settings, the two
  approaches perform very similarly in 3 settings, and Design-GRTS
  outperforms Model-GRTS in 3 settings.
\end{enumerate}

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-4-2.pdf}

@@

Figure \ref{fig:figdata} shows that mercury concentration is
right-skewed, with most lakes having a low value of mercury
concentration but a few having a much higher concentration. Mercury
concentration exhibits some spatial correlation, with high mercury
concentrations in lakes in the northeast and north central United
States. Because we are considering these lakes to be our entire
population, we know that the realized mean mercury concentration is
103.03 ng / g.

\begin{longtable}[]{@{}lrrrrr@{}}
\caption{\label{tab:appliedtab} Table XXX. Application of design-based
and model-based approaches to the NLA data set on mercury
concentration.}\tabularnewline
\toprule
Approach & Realized Mean & Estimate & SE & 95\% LB & 95\% UB \\
\midrule
\endfirsthead
\toprule
Approach & Realized Mean & Estimate & SE & 95\% LB & 95\% UB \\
\midrule
\endhead
Design IRS & 103.2 & 112.7 & 8.8 & 95.4 & 129.9 \\
Model IRS & 103.2 & 110.5 & 7.9 & 95.0 & 125.9 \\
Design GRTS & 103.2 & 101.8 & 6.1 & 89.8 & 113.7 \\
Model GRTS & 103.2 & 102.3 & 5.9 & 90.8 & 113.9 \\
\bottomrule
\end{longtable}

@@

\leavevmode\hypertarget{ref-stevens2003variance}{}%
Stevens Jr, D.L., Olsen, A.R., 2003. Variance estimation for spatially
balanced samples of environmental resources. Environmetrics 14,
593--610.

\leavevmode\hypertarget{ref-stevens2004spatially}{}%
Stevens Jr, D.L., Olsen, A.R., 2004. Spatially balanced sampling of
natural resources. Journal of the american Statistical association 99,
262--278.

>>>>

\begin{figure}
\includegraphics[width=1\linewidth]{SpatialDVM_Manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and two random samples. In the bottom row, we have two realizations of the same spatial process sampled at the same locations.}\label{fig:fig1}
\end{figure}

@@

Lisa M.: Need a more precise definition of ``miniature'' in this
context, and need an example.

The design-based approach can use spatial locations to obtain spatially
balanced samples. First we discuss spatial balance with respect to the
population (Stevens and Olsen, 2004). A sample is spatially balanced
with respect to the population if the sampled population units are a
miniature of the population units. A sample is a miniature of the
population if the distribution of the sampled population units mirrors
the density of all population units. Spatial balance with respect to the
population is different than spatial balance with respect to geography.
A sample that is spatially balanced with respect to geography is spread
out in some type of equidistant manner over geographical space and is
not meant to be miniatures of the population. When we refer to spatial
balance henceforth, we mean spatial balance with respect to the
population.

Spatially balanced samples are useful because they tend to yield
estimates that have lower variance than estimates constructed from
sampling designs lacking spatial balance (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens and Olsen, 2004; Wang et al., 2013). To quantify spatial
balance, Stevens and Olsen (2004) proposed loss functions based on
Voroni polygons. The first spatially balanced sampling algorithm that
saw widespread use was the Generalized Random Tessellation Stratified
(Stevens and Olsen, 2004). Since GRTS was developed, several other
spatially balanced sampling algorithms have emerged, including the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018). We focus on the Generalized Random
Tessellation Stratified (GRTS) algorithm to select spatially balanced
sampling because it has several attractive properties, including Lisa
M.: List major attractive properties, and detailed by Stevens and Olsen
(2004) and Dumelle et al. (2021).

The GRTS algorithm is used to sample from finite and infinite
populations and works by utilizing a mapping between two-dimensional and
one-dimensional space. The population units in two-dimensional space are
divided into cells using a hierarchical index. Population units are then
mapped to a one-dimensional line via the hierarchical indexing. The line
length of each population unit equals its inclusion probability. A
systematic sample is conducted on the line and these samples are linked
to a population unit in two-dimensional space, which results in the
desired sample. Stevens and Olsen (2004) and Dumelle et al. (2021)
provide further details.

After collecting a sample using the GRTS algorithm, the data are used to
estimate population parameters. The Horvitz-Thompson estimator (Horvitz
and Thompson, 1952) yields unbiased estimates of population means and
totals. For example, if \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\) (denoted by \(\hat{\tau}_{ht}\)),
is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) and \(\pi_i\) are the observed value and
inclusion probability of the \(i\)th population unit selected in the
sample. A similar formula exists for estimating the mean, \(\mu\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but they have two drawbacks. First, they rely
on calculating \(\pi_{ij}\), the probability that population unit \(i\)
and population unit \(j\) are included in the sample, and this can be
very difficult to calculate. Second, they ignore the spatial locations
of the population units. To address these drawbacks, Stevens and Olsen
(2003) proposed a local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\), and it
incorporates spatial locations by assigning higher weights to nearby
observations. Stevens and Olsen (2003) show this variance estimators
tends to reduce the estimated standard error of \(\hat{\tau}\), yielding
narrower confidence confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of basing inference off of a specific
sampling design, we assume the data are generated by a spatial process.
Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. Let
\(\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}\)
be a response variable that can be measured at the \(N\) population
units and is represented as an \(N \times 1\) vector. Suppose we want to
predict some linear function of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights. For
example, if we want to predict the population total across all
population units, then we would use a vector of 1's for the weights.

However, we often only have a sample of the \(N\) population units.
Denoting quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with a subscript \emph{u},

@@

We also assume that there is spatial correlation in \(\bm{\delta}\),
which can be modeled using a covariance function. It is common to assume
the covariance function is second-order stationary and isotropic
(Cressie, 1993), and that the spatial covariance decreases as the
separation between population units increases. Many spatial covariance
functions exist, but the primary function we use throughout the
simulations and applications in this manuscript is the exponential
covariance function: the \(i,j^{th}\) entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align} where \(h_{i,j}\) is the distance between population units
\(i\) and \(j\), and \(\bm{\theta}\) is a vector of spatial covariance
parameters of the partial sill \(\theta_1\), the range \(\theta_2\), and
the nugget \(\theta_3\); and, \(\mathbbm{1}\) is equal to 1 when
distance \(h_i,j\) is equal to 0, and equal to 0 otherwise. However, any
spatial covariance function could be used in the place of the
exponential, including functions that allow for non-stationarity or
anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

Lisa M. : Include formulas. Perhaps, but, these are very heavy in
notation and matrix algebra. We might consider, however, adding the
formulas to an Appendix.

@@

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. However,
k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013),
random forest (Breiman, 2001), Bayesian models (Chan-Golston et al.,
2020), among others, can also be used to obtain predictions for a mean
or total from spatially correlated responses in a finite population
setting. We choose to use FPBK because it is faster than a Bayesian
approach and random forest and because Ver Hoef and Temesgen (2013)
showed that the method outperforms k-nearest-neighbors in many
scenarios.

@@

  \begin{itemize}
  \tightlist
  \item
    normal: mean 0, variance 2
  \item
    lognormal: log scale mean 0, log scale variance 2 (total variance
    47)
  \item
    lognormalbig: log scale mean 0, log scale variance 4 (total variance
    2,926)
  \end{itemize}
\item
  sample sizes: \(n = 10, 50, 150; N = 900\)
\item
  layout: gridded vs random uniform population locations confined to a 1
  x 1 unit square
\end{itemize}

@@

There were 2000 trials for each simulation. The original response
(before exponentiating if applicable) for the dependent error cases was
normally distributed with an exponential covariance function with
partial sill of 0.9, effective range of \(\sqrt{2}\), and a nugget of
0.1. For the independent error cases, the partial sill was 0 and the
nugget was 1.

Lisa M. Notes: adding an intermediate level of spatial dependency?
Transfer simulation scenarios to a table Explain what the effective
range is Fill in the details of what exactly each approach means
(perhaps a table would be a good way to do this) Reorder the sims in the
first figure by some criterion. Think about what would be a ``reasonable
sample size'' instead of 10. Define medae If the data has a large
right-skew, wouldn't one consider a transformation before the analysis?
We should address this by stating that the BLUP for the log response
does not mean that e\^{}logBLUP is the BLUP for the response on the
original scale.

In each simulated data set, a GRTS sample and an IRS sample were
selected. Then for the GRTS sample, the design-based approach using the
local neighborhood variance (Design GRTS) and a model-based approach
were applied (Model GRTS). Then for the IRS sample, the design-based
approach using the simple random sample variance (Design IRS) and a
model-based approach were applied (Model IRS).

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  We will now focus on comparing Design-GRTS to Model-GRTS, the two best
  approaches for any reasonable sample size. In the independent error
  settings, the two approaches perform very similarly, so those results
  are omitted in the following graph. In the dependent error settings,
  using rmspe as the performance criterion, Model-GRTS outperforms
  Design-GRTS in 12 of the 18 settings, the two approaches perform very
  similarly in 3 settings, and Design-GRTS outperforms Model-GRTS in 3
  settings.
\end{enumerate}

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-5-2.pdf}

@@

Figure \ref{fig:figdata} shows that mercury concentration is
right-skewed, with most lakes having a low value of mercury
concentration but a few having a much higher concentration. Mercury
concentration exhibits some spatial correlation, with high mercury
concentrations in lakes in the northeast and north central United
States. Because we are considering these lakes to be our entire
population, we know that the realized mean mercury concentration is
103.2 ng / g.

\begin{longtable}[]{@{}lrrrr@{}}
\caption{\label{tab:appliedtab} Application of design-based and
model-based approaches to the NLA data set on mercury concentration. The
true mean concentration is 103.2 103.2 ng / g}\tabularnewline
\toprule
Approach & Estimate & SE & 95\% LB & 95\% UB \\
\midrule
\endfirsthead
\toprule
Approach & Estimate & SE & 95\% LB & 95\% UB \\
\midrule
\endhead
Design IRS & 112.7 & 8.8 & 95.4 & 129.9 \\
Model IRS & 110.5 & 7.9 & 95.0 & 125.9 \\
Design GRTS & 101.8 & 6.1 & 89.8 & 113.7 \\
Model GRTS & 102.3 & 5.9 & 90.8 & 113.9 \\
\bottomrule
\end{longtable}

@@

\leavevmode\hypertarget{ref-stevens2003variance}{}%
Stevens, D.L., Olsen, A.R., 2003. Variance estimation for spatially
balanced samples of environmental resources. Environmetrics 14,
593--610.

\leavevmode\hypertarget{ref-stevens2004spatially}{}%
Stevens, D.L., Olsen, A.R., 2004. Spatially balanced sampling of natural
resources. Journal of the american Statistical association 99, 262--278.
